---
title: "Benford's Law"
subtitle: "Probability Distributions with R"
author: "StatsResource: Specal Topics In Probabilty"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{css my-header-colors, echo = FALSE}

.page-header {
    background-image: linear-gradient(120deg, lightblue, steelblue);
    color: black;
}

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(VGAM)
```


## Power Laws

**Power-law distributions** describe patterns where small occurrences are extremely common, while large instances are rare but significantly impactful. These distributions follow the general form:

\[
P(x) \propto x^{-\alpha}
\]

where \( \alpha \) is a constant typically greater than 1.

Power laws appear across a wide range of both manmade and natural phenomena, such as:
- **City populations** (e.g. Tokyo vs. smaller towns)
- **Income and wealth** (Pareto Principle or 80/20 Rule)
- **Word frequencies** in human language
- **Internet traffic and website popularity**
- **Earthquake magnitudes**
- **Network connectivity** (social media, power grids, etc.)

### Notable Power-Law-Related Laws:
- **Zipf’s Law**: In a given body of text, the frequency of any word is inversely proportional to its rank.
- **Benford’s Law**: Predicts the frequency of digits in naturally occurring numerical datasets.
- **Bradford’s Law**: Describes how articles on a given topic are distributed across journals.

---

## Benford's Law

**Benford's Law**, also known as the *First-Digit Law*, reveals that in many real-world datasets, lower digits occur as the leading digit more frequently than higher ones.

### Leading Digit Probabilities:
- Digit **1**: ~**30.1%**
- Digit **2**: ~**17.6%**
- ...
- Digit **9**: ~**4.6%**

This distribution contrasts sharply with the intuition that all digits 1–9 should appear equally often (~11.1% each).

### Why It Works:
Benford’s Law holds especially well for datasets that:
- Span **several orders of magnitude**
- Are **not bounded artificially**
- Emerge from **multiplicative processes**, such as growth rates or natural measurements

It applies across fields—physics, economics, social sciences, and even biology—whenever these conditions are met.

### Historical Background:
- First noted by **Simon Newcomb** in 1881 when he observed that logarithmic tables' earlier pages were more worn.
- Later formalized and empirically tested by **Frank Benford** in 1938 using over 20 datasets.

---

## Applications of Benford's Law

Benford's Law is a powerful diagnostic tool for detecting unnatural patterns and suspicious data. It's especially valuable in:
- **Forensic accounting** and **financial audits**
- **Tax fraud detection**
- **Scientific data integrity checks**
- **Election result validation**
- **Ecological and environmental reporting**

When fabricated numbers or manipulated datasets diverge from the expected Benford distribution, red flags are raised for further investigation.

---

## Benford Distribution in R

Here’s how to compute and visualize the Benford distribution in R using the `VGAM` package:

```{r}
# Load the required library
library(VGAM)

# Compute probabilities for digits 1 through 9
benford_probs <- dbenf(1:9)

# Visualize the distribution
barplot(benford_probs,
        names.arg = 1:9,
        col = "steelblue",
        main = "Benford's First Digit Distribution",
        xlab = "Leading Digit",
        ylab = "Probability")
```
