---
title: "Testing the Assumption of Normality"
subtitle: "Statistics with R"
author: DragonflyStats.github.io
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)
```

Testing The Assumption of Normality
========================================

For example, a fundamental assumption of linear models (i.e. regression models) is that the residuals (differences between observed and predicted value) are normally distributed with mean zero.


The null hypothesis of both the `Anderson-Darling' and `Shapiro-Wilk' tests is that the population is normally distributed, and the alternative hypothesis is that the data is not normally distributed.

For both tests, the null and alternative hypothesis are :
\qquad $H_0 : $ The data set is normally distributed.
\qquad $H_1 : $ The data set is \textbf{not} normally distributed.

### Anderson-Darling Test
To implement the Anderson-Darling Test for Normality, one must first install the ***nortest*** package.


```{r}
library(nortest)
#Generate 100 normally distributed random numbers
NormDat = rnorm(100)
ad.test(NormDat)
```

### Shapiro-Wilk Test
The Shapiro-Wilk test is directly implementable, without loading any additional packages.


```{r}
#Generate 100 normally distributed random numbers

NormDat = rnorm(100)

shapiro.test(NormDat)
```

Sample output, using the randomly generated \texttt{NormDat} data set, is as follows:
```{r}
shapiro.test(NormDat)

```
Here, the p-value is well above the 0.05 threshold. Hence we \textbf{fail to reject} the null hypothesis, and may proceed to treat the \texttt{NormDat} data set as normally distributed.

### Graphical Procedures for Assessing Normality}

There are two useful graphical methods for determining whether a data set was normally distributed. The first is the histogram, which we have seen previously. If the histogram is reasonably bell-shaped, then the data can be assumed to be normally distributed. The relevant R command is ***hist()***.


The second is the ***quantile-quantile plot*** (or QQ-plot).
For assessing normality, we implement a qq-plot  using the ***qqnorm()*** function.

Additionally the command ***qqline()*** function adds a trendline to a normal quantile-quantile plot. If the data is normally distributed, then the points on the plot follow the trendline.


```{r}
#Generate 100 normally distributed random numbers

NormDat = rnorm(100)

qqnorm(NormDat)
qqline(NormDat)
```



### Transforming the Data

Sometimes when we get non-normal data, we can change the scale of our data i.e. transform it to get a normal distribution. One transformation that often works for positively skewed data is the natural logarithm (ln) transformation.

In such a case, we work with the natural logarithms of the data set, rather than the data itself.